Hamiltonian mechanics is conservation of information entropy

### 2018-10-28 Articles/HamiltonianMechanicsIsConservationOfEntropy/article.tex
Added Christine's comments

\title{Hamiltonian mechanics is conservation of information entropy}
\author{Christine A. Aidala, Gabriele Carcassi}

@@

\begin{abstract}
	In this work we show the equivalence between Hamiltonian mechanics and conservation of information entropy. We will show that distribution with coordinate independent values for information entropy require that the manifold on which the distribution is defined is charted by conjugate pairs (i.e. it is a symplectic manifold). We will also show that further requiring that the information entropy is conserved during the evolution yields Hamilton's equations. This will provide new insights in the nature of classical mechanics and physics in general, as it will connect to elements of relativity, statistical mechanics, thermodynamics, quantum mechanics, probability theory and symplectic geometry. This work is part of a larger project, called Assumptions of physics, which aims to rederive the known equations from a handful of physically meaningful starting points.
\end{abstract}

@@

Over the past decade there has been a renewed interest in the underpinning of classical mechanics~\cite{North,Curiel,Barrett1,Barrett2}. Different attempts and arguments have been put forward to understand what are the most important feature of the different mathematical formulations, whether one is more fundamental of the other in some sense or whether they are equivalent.

@@

It should be clear that we do not fault the authors. Even just a century ago, there was a push to find fundamental principles or laws that could serve as a starting point for the different theories. Newtonian mechanics, thermodynamics and special relativity serve as good examples. Nowadays, unfortunately, much of the current fundamental theories, including Lagrangian, Hamiltonian, Quantum mechanics and Quantum Field Theory take as their starting point a particular mathematical structure. That is, we do not know what the value of a Lagrangian is or what the symplectic form represents: we just take them as a given. Moreover, a lot of work in theoretical physics nowadays starts by exploring novel abstract mathematical structures and then trying to see whether some physical meaning or prediction can be extracted. We are not going to debate here how and why this shift happened. We just merely note that it leaves the semantics of our theories ill defined.\footnote{Which is a nice way to say that we don't really know what are we talking about.}

It is only natural that, given the circumstances, one uses the only thing that seems to be well defined, the math, and tries to get understanding from there. But here lies the fundamental problem: that math is not created by physicists to do physics, it is created by mathematicians to do math. The definitions of topology, differentiable manifold, symplectic manifold, Hilbert space and so on are chosen either because they are convenient to prove theorems or because they are convenient to make calculations.\footnote{In fact, it is common in math to have two such definitions and then prove they are equivalent.} We should not expect them to be a good match for definitions that would be physically meaningful. In other words, there is a very good chance that just looking at just the math may lead us, like horses with blinders, to a very clear path which is not the physically meaningful one.\footnote{To be clear, we do not fault the mathematicians either as they are not responsible for whether their structures are used appropriately or not.}

Fully conscious of these issues, we have embarked in an effort, called Assumptions of Physics~\cite{carc1}, that aims to find a handful of physical principles from which the known laws of physics can be formally derived~\cite{carc2}. The approach has several advantages. First, by starting from physical ideas we put physics back at the center of the discussion, and not relegate it as a mere after the fact interpretation: the semantics is always well defined. Second, it puts physics on sturdier mathematical grounds as we have a precise physical meaning to all the formal objects and we leave out those that do not. Third, it fosters connections between different fields of science, math, philosophy and engineering: it is our belief that nature is one and does not care about such divisions and therefore truly fundamental concepts must span freely across different areas. Fourth, it gives a framework to pose deep questions of the type the philosophy of science community is interested in. And lastly, by being forced to formalize and make explicit our assumptions, we can better understand how and why they fail possibly giving new insights that may lead to new physics. While the project is ambitious, we have already made progress beyond what we thought it could be reasonably done. So there's that.

In this paper we focus on just one particular insight provided by our work: \textbf{a distribution over a manifold conserves information entropy over time if and only if its elements evolve according to Hamiltonian mechanics.} That is, the single requirement of information entropy conservation not only gives us Hamiltonian dynamics but also the structure of phase space (i.e. a symplectic structure where variables are organized in conjugate pairs). In section 2 we show that, in general, information entropy and distributions densities are coordinate dependent. In section 3 we show that the structure of phase space is required to make such objects invariant under coordinate transformation (i.e. observer independent) and therefore give them an objective physical character. Conservation of entropy over time will force that invariance during time evolution, giving rise to Hamilton's equations. In chapter 4 we will briefly discuss the result and show how it necessarily leads to a classical version of the uncertainty principle.

The overall idea is that \textbf{Hamiltonian mechanics is best understood as the deterministic and reversible\footnote{By reversible we mean the ability, given a state, to reconstruct its past. This is the dual of determinism, that is the ability, given a state, to predict its future.} evolution of infinitesimally reducible objects}. As the information needed to identify an element of the distribution does not change in time, knowing the initial state is equivalent to knowing the future or past state. Classical particles, then, should be understood as the limit of subsequent subdivisions, infinitesimally small objects with extent in phase space. This single physical model is captured piecemeal by elements of statistics, probability theory and symplectic geometry which are given a coherent and well defined meaning. In our view, this understanding also meshes better with other ideas from general relativity\footnote{A point particle in general relativity would not follow a geodesic and would, due to the infinite mass density, puncture space-time.}, thermodynamics and quantum mechanics.

@@

While the mathematical setting is the same for degrees of belief (credence) or the likelyhood that something will happen (probability), we will assume we are working with an actual physical distribution. Even when we are talking about marginal distributions or information entropy, we are doing it in the context of a system that has parts and each part has a different characterization.

The link to information theory is the following: suppose we have a distribution, known to both of us, and you select one element at random. How many yes or no questions do you have to answer before I know what case you selected? For example, suppose there are 4 balls of which 2 are red and 2 are green, one yes or no question would be sufficient. In general, if $I$ is the number of questions then $C = 2 ^ I$ are the number of cases you can distinguish with those questions. On the other hand, suppose $\rho(\xi_0)=1/5$, then we know that we'd select $\xi_0$ one in 5 cases. That is, the normalized distribution is the inverse of the number of cases: $\rho(\xi_0) = \frac{1}{C}$.  Therefore the information needed to identify a particular element is $I(\xi^a)=\log \frac{1}{\rho(\xi^a)}$. If we take the average and change the base of the logarithm, we have:
\begin{equation}
I(\rho(\xi^a)) = \int \rho(\xi^a)) \ln \left(\frac{1}{\rho(\xi^a))}\right) d\xi^1 d\xi^2 ... d\xi^n =-\int \rho(\xi^a)) \ln (\rho(\xi^a)) d\xi^1 d\xi^2 ... d\xi^n
\end{equation}
which is the information entropy of the distribution.\footnote{The nature of information entropy is often misunderstood, especially in physics. It does not represent the information the distribution has: it represent the \emph{additional} information that would be needed to identify an element. Also, it does not represent knowledge: as we are studying actual physical distributions, any correct distribution has full knowledge of the system.}

Technically, given the more convenient choice of the natural logarithm, the unit of information is the nat, which is equal to $\frac{1}{\ln 2}$ bits. However, in the discussion we will refer to it as measured in bits, as we believe that the majority of the readers is more conformable with the latter unit. Note that for continuous variables the expression has a slightly different meaning: the uniform distribution in a unit volume is assigned zero entropy so $I(\rho(\xi^a))$ represents the relative number of bits to achieve the same level of precision, which can be negative.

@@

There is one feature about densities we need to be fully aware of: they are not invariant under change of variables. Intuitively, if we change units of the denominator the distribution changes value. For example, $1$ kg/m becomes $1000$ kg/km: the number changes and the units allow us to understand that the overall quantity remains the same but mathematics does not keep track of units. In general, if $\hat{\xi}^a=\hat{\xi}^a(\xi^b)$ we can recover how the density changes by requiring that the total in each volume is the same no matter what variables are used. That is:
\begin{align*}
\int_V \rho(\xi^b) d\xi^1 ... d\xi^n &= \int_V \rho(\hat{\xi}^a) d\hat{\xi}^1 ... d\hat{\xi}^n \\
&=\int_V\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\xi^1 ... d\xi^n
\end{align*}
\begin{equation}\label{density_transformation}
\rho(\xi^b) = \rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|
\end{equation}
the density is multiplied by the absolute value of the Jacobian determinant of the transformation. As the Jacobian is in general a function of the variables themselves, this also means that whether the density at one point is greater or smaller than at another also depends on the choice of variables.

Similarly, we can see that information entropy is not invariant
\begin{align*}
I(\rho(\xi^b)) &=-\int \rho(\xi^b) \ln (\rho(\xi^b)) d\xi^1 ... d\xi^n \\
&=-\int \rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| \ln \left(\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|\right) d\xi^1 ... d\xi^n \\
&=-\int \rho(\hat{\xi}^a) \ln \left(\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|\right) d\hat{\xi}^1 ... d\hat{\xi}^n \\
&=-\int \rho(\hat{\xi}^a) \ln (\rho(\hat{\xi}^a)) d\hat{\xi}^1 ... d\hat{\xi}^n -\int \rho(\hat{\xi}^a) \ln \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\hat{\xi}^1 ... d\hat{\xi}^n
\end{align*}
\begin{equation}\label{entropy_transformation}
I(\rho(\xi^b)) =I(\rho(\hat{\xi}^a)) -\int \rho(\hat{\xi}^a) \ln \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\hat{\xi}^1 ... d\hat{\xi}^n
\end{equation}
The reason is that we may be changing scale and therefore level of precision. Given that both change is determined by the Jacobian, we conclude that \textbf{during a change of variable the information entropy of all possible distributions does not change if and only if the density remains the same at every point under the same change of variable}.

@@

In probability theory one starts with three objects: a set of outcomes $\mathcal{M}$, a set of events $\sigma_\mathcal{M}$ where each event is a collection of outcomes, and a probability measure $\mu : \sigma_\mathcal{M} \rightarrow [0,1]$ that assigns a number between zero and one to each event. That is, we don't assign a probability density to the points of $\mathcal{M}$ but we assign a finite probability to finite regions. We can then define a random variable $\xi : \mathcal{M} \rightarrow \mathbb{R}$ as a real valued function of the outcomes. For each random variable we can define its cumulative distribution function $F_\xi(\xi_0)=\mu(\xi<\xi_0)$ and the probability density function is its Radon-–Nikodym derivative $\rho(\xi) = \frac{dF_\xi}{d\xi}$. That is: the integral is the primal object while the density is derived.

This approach actually makes more physical sense: what we measure is the amount of mass in a finite region and the density is the limit for a smaller and smaller ones. The fundamental objects (outcomes, events and probability) are all defined without reference to variables. But in probability theory we have no notion of coordinate systems, vectors and all other geometrical notions.

In differential geometry, one starts with a manifold: a set of points $\mathcal{M}$ which can be given local coordinates $\xi^i : \mathcal{M} \to \mathbb{R}$. One defines a tangent space $\mathsf{T}\mathcal{M}$ where vectors live and cotangent space $\mathsf{T}^*\mathcal{M}$ were linear functions of vectors live. Then we define $n$-forms as multi-linear functions of $n$ vectors that return the value associated with the infinitesimal parallelepiped they form. That is $\nu : (\mathsf{T}^*\mathcal{M})^n \rightarrow \mathbb{R}$. As vectors are coordinate invariant, $n$-forms are coordinate invariant as well so we can write $\mu(V)=\int_V \nu(dV)$ with no reference to coordinates. If $e_a$ are the basis vector associated with $\xi^a$ and $e^b$ are linear functions such that $e^b(e_a)=\delta_a^b$, then we can express each $n$-form as $\rho(\xi^a)e_1\wedge e_2 \wedge ... \wedge e_n$. The density $\rho(\xi^a)$ is the component of the $n$-form expressed in the $\xi^a$ coordinates in the a way analogous to components of a vector.\footnote{Technically, the form is linear so it changes with the Jacobian determinant and not its absolute value. But negative determinants means changing orientation of the region of integration, so fixing that would bring another sign change and $\rho(\xi^a)$ would not change. This is equivalent to saying that some coordinate systems are right handed and some are left handed.}

This approach gives us a way to understand which objects are coordinate independent and which aren't. But it has two issues: it does not allow to easily talk about coordinate dependent operations, such as marginal distributions and information entropy, and it may use definitions that do not map well to physical concepts.\footnote{Typically, a vector is defined as a map between a scalar function of the manifold to another scalar function to the manifold. Or as an equivalence class of trajectories. It is a stretch to think of fluid velocities or electric fields in those terms. For that same reason, we do not use the differential geometry notation $\frac{\partial}{\partial \xi^a}$ and $d\xi^a$ for vector and covector basis because they do not convey the actual physical meaning of these objects as they are used in physics.}

So the situation is that we have a single clear physical objects we want to study, but different and somewhat disconnected mathematical frameworks to study it. This should not be surprising as the frameworks are defined by mathematicians, as we said before, to solve their own needs. But it is an issue when these are used in physics without careful consideration as each mathematical framework only partially captures our physical understanding. That is why we argue that simply examining the mathematical structure used in a physical theory is not enough to gain a complete picture.

@@

Suppose we have a system which we can think of made of infinitesimally small parts, which we call particles. Suppose  that each infinitesimal part can be assigned a state from a state space $\mathcal{S}$. The state of the whole system will be a distribution over those states. Now suppose that this system evolves in a way that is deterministic and reversible, that is the state of a particle at one time is associated with one and only one state at another time. What does that tell us about our distribution?

If the evolution is deterministic and reversible, then all the parts that are in one state will be mapped to another: the density should be exactly mapped to a different state and therefore should not change value. If the evolution is deterministic and reversible, then if I give you enough information to identify the state of an element at one time I also gave you enough information to identify the state of the same element at a future time: the information entropy will remain the same over time. And here lies the problem: how can we make sure that these quantities are preserved over time if they are not even observer independent? How can we know if the density stays the same if we can't even compare the value of the densities at two different elements of the distribution? How can we tell whether the information entropy is conserved if we can't event define a unique value at a given time?

This leads us to the following insight: \textbf{if we want to define deterministc and reversible evolution over distributions, they must be invariant distributions. That is, the densities and information entropy must be independent on the choice of coordinates.} As we noted before, requiring one to be coordinate invariant will automatically require the other to be invariant as well.

If our variables are discrete, all our integrals are sums and all our definitions work. Everything is already coordinate invariant. If our variables are continuous, instead, we have a problem. Because the densities and the information entropy vary with the Jacobian, we need something more. Our state space $\mathcal{S}$ must allow to write densities whose value does not change under coordinate transformation. This seems to be impossible given what we saw before: how can it work?

@@

If we have a set of variables fixed by a single choice of unit then what we discussed before is the only possible way. Let $q$ be the coordinate, the variable that defines the unit. Suppose we have $n$ variable $\xi^a$ such that $\xi^1 = q$. Each vector $v^a$ tangent to that space has $n$-components. If we change unit $\hat{q}=\hat{q}(q)$ we have two constraints: the one unit change and the unitary Jacobian, and these must be enough to determine how the vector components change. The components must be equal to the constraints and therefore they are exactly two variable $\xi^a = (q, k)$. Then $k$ is the conjungate variable of $q$.

@@

Let's set $\hbar$ as the unit we use to describe the amount of possible states (or the inverse of a unitary density). We have:
\begin{equation}
m_V = M \mu(V) =\int_V M \rho(q, k) \hbar dq dk = \int_V M \rho(q, p) dq dp
\end{equation}
where we set $p=\hbar k$. We should recognize $p$ as conjugate momentum and $k$ the classical analogue of the wave number. Suppose we change variables. In general the Jacobian will be given by:
\begin{equation}
\label{Poisson}
\begin{aligned}
|J| &= \left| \begin{matrix}
\dfrac{\partial \hat{q}}{\partial q} & \dfrac{\partial \hat{q}}{\partial p} \\[2.2ex]
\dfrac{\partial \hat{p}}{\partial q} & \dfrac{\partial \hat{p}}{\partial p} \end{matrix} \right| = \frac{\partial \hat{q}}{\partial q} \frac{\partial \hat{p}}{\partial p} - \frac{\partial \hat{p}}{\partial x} \frac{\partial \hat{q}}{\partial p} &= \{\hat{q}, \hat{p}\}
\end{aligned}
\end{equation}
which we recognize it to be the Poisson bracket. If the change of variable is simply a change of coordinates, a change of units, we have:
\begin{equation}
\label{coordinate_change}
\begin{aligned}
\hat{q} &= \hat{q}(q) \\
\{\hat{q}, \hat{p}\} &= 1 = \frac{\partial \hat{q}}{\partial q} \frac{\partial \hat{p}}{\partial p} \\
\dfrac{\partial \hat{p}}{\partial p} &= \frac{\partial \hat{q}}{\partial q} ^{-1} = \frac{\partial q}{\partial \hat{q}} \\
\hat{p} &= \frac{\partial q}{\partial \hat{q}} p
\end{aligned}
\end{equation}
Conjugate momentum transforms as a covariant component. Under coordinate transformations we have:
\begin{equation}
\label{density_invariance}
\begin{aligned}
\rho(q,p) &= \rho(\hat{q}, \hat{p}) \\
I(\rho(q,p)) &= I(\rho(\hat{q},\hat{p}))
\end{aligned}
\end{equation}
Let us call $(q,p)$ canonical variables if the density is equal to the invariant density. Then two other variables such that $\{\hat{q}, \hat{p}\}=1$ are also canonical since the Jacobian determinant will be unitary.

>>>>

\title{Hamiltonian mechanics is conservation of information entropy}
\author{Gabriele Carcassi, Christine A. Aidala}

@@

\begin{abstract}
	In this work we show the equivalence between Hamiltonian mechanics and conservation of information entropy. We will show that distributions with coordinate independent values for information entropy require that the manifold on which the distribution is defined is charted by conjugate pairs (i.e. it is a symplectic manifold). We will also show that further requiring that the information entropy is conserved during the evolution yields Hamilton's equations. This will provide new insights into the nature of classical mechanics and physics in general, as it will connect to elements of relativity, statistical mechanics, thermodynamics, quantum mechanics, probability theory and symplectic geometry. This work is part of a larger project, called Assumptions of Physics, which aims to rederive the known equations of physics from a handful of physically meaningful starting points.
\end{abstract}

@@

Over the past decade there has been a renewed interest in the underpinning of classical mechanics~\cite{North,Curiel,Barrett1,Barrett2}. Different attempts and arguments have been put forward to understand what are the most important features of the different mathematical formulations, whether one is more fundamental than the other in some sense or whether they are equivalent.

@@

It should be clear that we do not fault the authors. Even just a century ago, there was a push to find fundamental principles or laws that could serve as a starting point for the different theories. Newtonian mechanics, thermodynamics and special relativity serve as good examples. Nowadays, unfortunately, many of the current fundamental theories, including Lagrangian, Hamiltonian, quantum mechanics and quantum field theory take as their starting point a particular mathematical structure. That is, we do not know what the value of a Lagrangian means or what the symplectic form represents: we just take them as given. Moreover, a lot of work in theoretical physics nowadays starts by exploring novel abstract mathematical structures and then trying to see whether some physical meaning or prediction can be extracted. We are not going to debate here how and why this shift happened. We merely note that it leaves the semantics of our theories ill defined.\footnote{Which is a nice way to say that we don't really know what we are talking about.}

It is only natural that, given the circumstances, one uses the only thing that seems to be well defined, the math, and tries to get understanding from there. But here lies the fundamental problem: that math is not created by physicists to do physics, it is created by mathematicians to do math. The definitions of topology, differentiable manifold, symplectic manifold, Hilbert space and so on are chosen either because they are convenient to prove theorems or because they are convenient to make calculations.\footnote{In fact, it is common in math to have two such definitions and then prove they are equivalent.} We should not expect them to be a good match for definitions that would be physically meaningful. In other words, there is a very good chance that looking at just the math may lead us, like horses with blinders, to a very clear path which is not the physically meaningful one.\footnote{To be clear, we do not fault the mathematicians either as they are not responsible for whether their structures are used appropriately or not.}

Fully conscious of these issues, we have embarked nn an effort, called Assumptions of Physics~\cite{carc1}, that aims to find a handful of physical principles from which the known laws of physics can be formally derived~\cite{carc2}. The approach has several advantages. First, by starting from physical ideas we put physics back at the center of the discussion, instead of relegating it to a mere after-the-fact interpretation: the semantics is always well defined. Second, it puts physics on sturdier mathematical grounds as we have a precise physical meaning to all the formal objects and we leave out those that do not. Third, it fosters connections between different fields of science, math, philosophy and engineering: nature is one and does not care about such divisions and therefore truly fundamental concepts must span freely across different areas. Fourth, it gives a framework to pose questions of the type in which the philosophy of science community is interested. And lastly, by being forced to formalize and make explicit our assumptions, we can better understand how and why they fail, possibly giving new insights that may lead to new physics. While the project is ambitious, we have already made progress beyond what we thought could be reasonably done.

In this paper we focus on just one particular insight provided by our work: \textbf{a distribution over a manifold conserves information entropy over time if and only if its elements evolve according to Hamiltonian mechanics.} That is, the single requirement of information entropy conservation not only gives us Hamiltonian dynamics but also the structure of phase space (i.e. a symplectic structure where variables are organized in conjugate pairs). In section 2 we show that, in general, information entropy and densities are coordinate dependent. In section 3 we show that the structure of phase space is required to make such objects invariant under coordinate transformation (i.e. observer independent) and therefore give them an objective physical character. Conservation of entropy over time will enforce that invariance during time evolution, giving rise to Hamilton's equations. In section 4 we will briefly discuss the result and show how it necessarily leads to a classical version of the uncertainty principle.

The overall idea is that \textbf{Hamiltonian mechanics is best understood as the deterministic and reversible\footnote{By reversible we mean the ability, given a state, to reconstruct its past. This is the dual of determinism, that is the ability, given a state, to predict its future.} evolution of infinitesimally reducible objects}. As the information needed to identify an element of the distribution does not change in time, knowing the current state is equivalent to knowing the future or past state. Classical particles, then, should be understood as the limit of subsequent subdivisions, infinitesimally small objects with extent in phase space. This single physical model is captured piecemeal by elements of statistics, probability theory and symplectic geometry which are now given a coherent and well defined meaning. In our view, this understanding also meshes better with other ideas from general relativity\footnote{A point particle in general relativity would not follow a geodesic and would, due to the infinite mass density, puncture space-time.}, thermodynamics and quantum mechanics.

@@

While the mathematical setting is the same for degrees of belief (credence) or the likelihood that something will happen (probability), we will assume we are working with an actual physical distribution. Even when we are talking about marginal distributions or information entropy, we are doing it in the context of a system that has parts and each part has a different characterization.

The link to information theory is the following: suppose we have a distribution, known to both of us, and you select one element at random. How many yes or no questions do you have to answer before I know what case you selected? For example, suppose there are 4 balls of which 2 are red and 2 are green, one yes or no question would be sufficient. In general, if $I$ is the number of questions then $C = 2 ^ I$ are the number of cases you can distinguish with those questions. On the other hand, suppose for a particular value $\xi_0$ of a discrete distribution we have $\rho(\xi_0)=1/5$, then we know that we'd select $\xi_0$ one in 5 cases. That is, the normalized distribution is the inverse of the number of cases: $\rho(\xi_0) = \frac{1}{C}$.  Therefore the information needed to identify a particular case is $I(\xi^a)=\log \frac{1}{\rho(\xi^a)}$. If we take the average and change the base of the logarithm, we have:
\begin{equation}
I(\rho(\xi^a)) = \int \rho(\xi^a) \ln \left(\frac{1}{\rho(\xi^a)}\right) d\xi^1 d\xi^2 ... d\xi^n =-\int \rho(\xi^a) \ln (\rho(\xi^a) d\xi^1 d\xi^2 ... d\xi^n
\end{equation}
which is the information entropy of the distribution.\footnote{The nature of information entropy is often misunderstood, especially in physics. It does not represents the information the distribution has: it represent the \emph{additional} information that would be needed to identify an element. Also, it does not represent knowledge: as we are studying actual physical distributions, any correct distribution represents full knowledge of the system.}

Technically, given the more convenient choice of the natural logarithm, the unit of information is the nat, which is equal to $\frac{1}{\ln 2}$ bits. However, in the discussion we will refer to it as measured in bits, as we believe that the majority of the readers is more comfortable with the latter unit. Note that for continuous variables the expression has a slightly different meaning: the uniform distribution in a unit volume is assigned zero entropy so $I(\rho(\xi^a))$ represents the relative number of bits to achieve the same level of precision, which can be negative.

@@

There is one feature about densities we need to be fully aware of: they are not invariant under change of variables. Intuitively, if we change units the density changes value. For example, $1$ kg/m becomes $1000$ kg/km: the number changes and the units allow us to understand that the overall quantity remains the same but mathematics does not keep track of units. In general, if $\hat{\xi}^a=\hat{\xi}^a(\xi^b)$ we can recover how the density changes by requiring that the total in each volume is the same no matter what variables are used. That is:
\begin{align*}
\int_V \rho(\xi^b) d\xi^1 ... d\xi^n &= \int_V \rho(\hat{\xi}^a) d\hat{\xi}^1 ... d\hat{\xi}^n \\
&=\int_V\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\xi^1 ... d\xi^n
\end{align*}
\begin{equation}\label{density_transformation}
\rho(\xi^b) = \rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|
\end{equation}
the density is multiplied by the absolute value of the Jacobian determinant of the transformation. As the Jacobian is in general a function of the variables themselves, this means that whether the density at one point is greater or smaller than at another also depends on the choice of variables.

Similarly, we can see that information entropy is not invariant
\begin{align*}
I(\rho(\xi^b)) &=-\int \rho(\xi^b) \ln (\rho(\xi^b)) d\xi^1 ... d\xi^n \\
&=-\int \rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| \ln \left(\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|\right) d\xi^1 ... d\xi^n \\
&=-\int \rho(\hat{\xi}^a) \ln \left(\rho(\hat{\xi}^a) \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\|\right) d\hat{\xi}^1 ... d\hat{\xi}^n \\
&=-\int \rho(\hat{\xi}^a) \ln (\rho(\hat{\xi}^a)) d\hat{\xi}^1 ... d\hat{\xi}^n -\int \rho(\hat{\xi}^a) \ln \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\hat{\xi}^1 ... d\hat{\xi}^n
\end{align*}
\begin{equation}\label{entropy_transformation}
I(\rho(\xi^b)) =I(\rho(\hat{\xi}^a)) -\int \rho(\hat{\xi}^a) \ln \left\|\frac{\partial \hat{\xi}^a}{\partial \xi^b}\right\| d\hat{\xi}^1 ... d\hat{\xi}^n
\end{equation}
The reason is that we may be changing scale and therefore level of precision. Given that the change in both density and entropy is determined by the Jacobian, we conclude that \textbf{during a change of variable the information entropy of all possible distributions does not change if and only if the density remains the same at every point under the same change of variable}.

@@

In probability theory one starts with three objects: a set of outcomes $\mathcal{M}$, a set of events $\sigma_\mathcal{M}$ where each event is a collection of outcomes, and a probability measure $\mu : \sigma_\mathcal{M} \rightarrow [0,1]$ that assigns a number between zero and one to each event. That is, we don't assign a probability density to the points of $\mathcal{M}$ but we assign a finite probability to finite regions. We can then define a random variable $\xi : \mathcal{M} \rightarrow \mathbb{R}$ as a real valued function of the outcomes. For each random variable we can define its cumulative distribution function $F_\xi(\xi_0)=\mu(\xi<\xi_0)$ and the probability density function is its Radon-–Nikodym derivative $\rho(\xi) = \frac{dF_\xi}{d\xi}$. That is, the integral is the primal object while the density is derived.

This approach actually makes more physical sense: what we measure is the amount of mass in a finite region and the density is the limit for a smaller and smaller one. The fundamental objects (outcomes, events and probability) are all defined without reference to variables. But in probability theory we have no notion of coordinate systems, vectors and all other geometrical notions.

In differential geometry, one starts with a manifold: a set of points $\mathcal{M}$ which can be given local coordinates $\xi^i : \mathcal{M} \to \mathbb{R}$. One defines a tangent space $\mathsf{T}\mathcal{M}$ where vectors live and cotangent space $\mathsf{T}^*\mathcal{M}$ were linear functions of vectors live. Then we define $n$-forms as multi-linear functions of $n$ vectors that return the value associated with the infinitesimal parallelepiped they form. That is $\nu : (\mathsf{T}\mathcal{M})^n \rightarrow \mathbb{R}$. As vectors are coordinate invariant, $n$-forms are coordinate invariant as well so we can write $\mu(V)=\int_V \nu(dV)$ with no reference to coordinates. If $e_a$ are the basis vectors associated with $\xi^a$ and $e^b$ are linear functions such that $e^b(e_a)=\delta_a^b$, then we can express each $n$-form as $\rho(\xi^a)e_1\wedge e_2 \wedge ... \wedge e_n$. The density $\rho(\xi^a)$ is the component of the $n$-form expressed in the $\xi^a$ coordinates in a way analogous to components of a vector.\footnote{Technically, the form is linear so it changes with the Jacobian determinant and not its absolute value. But a negative determinant means changing orientation of the region of integration, so fixing that would bring another sign change and $\rho(\xi^a)$ would not change. This is equivalent to saying that some coordinate systems are right handed and some are left handed.}

This approach gives us a way to understand which objects are coordinate independent and which aren't. But it has two issues: it does not allow us to easily talk about coordinate dependent operations, such as marginal distributions and information entropy, and it may use definitions that do not map well to physical concepts.\footnote{Typically, a vector is defined as a map between a scalar function of the manifold to another scalar function of the manifold, or as an equivalence class of trajectories. It is a stretch to think of fluid velocities or electric fields in those terms. For that same reason, we do not use the differential geometry notation $\frac{\partial}{\partial \xi^a}$ and $d\xi^a$ for vector and covector basis because they do not convey the actual physical meaning of these objects as they are used in physics.}

So the situation is that we have a single clear physical object we want to study, but different and somewhat disconnected mathematical frameworks to study it. This should not be surprising as the frameworks are defined by mathematicians, as we said before, to solve their own needs. But it is an issue when these are used in physics without careful consideration as each mathematical framework only partially captures our physical understanding. That is why we argue that simply examining the mathematical structure used in a physical theory is not enough to gain a complete picture.

@@

Suppose we have a system which we can think of as made of infinitesimally small parts, which we call particles. Suppose  that each infinitesimal part can be assigned a state from a state space $\mathcal{S}$. The state of the whole system will be a distribution over those states. Now suppose that this system evolves in a way that is deterministic and reversible, that is the state of a particle at one time is associated with one and only one state at another time. What does that tell us about our distribution?

If the evolution is deterministic and reversible, then all the parts that are in one state will be mapped to another: the density should be exactly mapped to a different state and therefore should not change value. If the evolution is deterministic and reversible, then if I give you enough information to identify the state of an element at one time I also gave you enough information to identify the state of the same element at a future time: the information entropy will remain the same over time. And here lies the problem: how can we make sure that these quantities are preserved over time if they are not even observer independent? How can we know if the density stays the same if we can't even compare the value of the densities at two different elements of the distribution? How can we tell whether the information entropy is conserved if we can't even define a unique value at a given time?

This leads us to the following insight: \textbf{if we want to define deterministic and reversible evolution over distributions, they must be invariant distributions. That is, the densities and information entropy must be independent of the choice of coordinates.} As we noted before, requiring one to be coordinate invariant will automatically require the other to be invariant as well.

If our variables are discrete, all our integrals are sums and all our definitions work. Everything is already coordinate invariant. If our variables are continuous, instead, we have a problem. Because the densities and the information entropy vary with the Jacobian, we need something more. Our state space $\mathcal{S}$ must allow us to write densities whose value does not change under coordinate transformation. This seems to be impossible given what we saw before: how can it work?

@@

If we have a set of variables fixed by a single choice of unit then what we discussed before is the only possible way. Let $q$ be the coordinate, the variable that defines the unit. Suppose we have $n$ variable $\xi^a$ such that $\xi^1 = q$. Each vector $v^a$ tangent to that space has $n$-components. If we change unit $\hat{q}=\hat{q}(q)$ we have two constraints: the one unit change and the unitary Jacobian, and these must be enough to determine how the vector components change. The number of components must be equal to the number of constraints and therefore there are exactly two variables $\xi^a = (q, k)$. Then $k$ is the conjungate variable of $q$.

@@

Let's set $\hbar$ as the unit we use to describe the amount of possible states (or the inverse of a unitary density). We have:
\begin{equation}
m_V = M \mu(V) =\int_V M \rho(q, k) \hbar dq dk = \int_V M \rho(q, p) dq dp
\end{equation}
where we set $p=\hbar k$. We should recognize $p$ as conjugate momentum and $k$ the classical analogue of the wave number. Suppose we change variables. In general the Jacobian will be given by:
\begin{equation}
\label{Poisson}
\begin{aligned}
|J| &= \left| \begin{matrix}
\dfrac{\partial \hat{q}}{\partial q} & \dfrac{\partial \hat{q}}{\partial p} \\[2.2ex]
\dfrac{\partial \hat{p}}{\partial q} & \dfrac{\partial \hat{p}}{\partial p} \end{matrix} \right| = \frac{\partial \hat{q}}{\partial q} \frac{\partial \hat{p}}{\partial p} - \frac{\partial \hat{p}}{\partial q} \frac{\partial \hat{q}}{\partial p} &= \{\hat{q}, \hat{p}\}
\end{aligned}
\end{equation}
which we recognize to be the Poisson bracket. If the change of variable is simply a change of coordinates, a change of units, we have:
\begin{equation}
\label{coordinate_change}
\begin{aligned}
\hat{q} &= \hat{q}(q) \\
\{\hat{q}, \hat{p}\} &= 1 = \frac{\partial \hat{q}}{\partial q} \frac{\partial \hat{p}}{\partial p} \\
\dfrac{\partial \hat{p}}{\partial p} &= \frac{\partial \hat{q}}{\partial q} ^{-1} = \frac{\partial q}{\partial \hat{q}} \\
\hat{p} &= \frac{\partial q}{\partial \hat{q}} p
\end{aligned}
\end{equation}
Conjugate momentum transforms as a covariant component. Under coordinate transformations we have:
\begin{equation}
\label{density_invariance}
\begin{aligned}
\rho(q,p) &= \rho(\hat{q}, \hat{p}) \\
I(\rho(q,p)) &= I(\rho(\hat{q},\hat{p}))
\end{aligned}
\end{equation}
Let us call $(q,p)$ canonical variables if the density is equal to the invariant density. Then two other variables such that $\{\hat{q}, \hat{p}\}=1$ are also canonical since the Jacobian determinant will be unitary.
